---
title: "COVID-19 Project - Simulation of States' Response to COVID-19"
author: "Jordan Lardieri"
date: "4/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# How does social distancing and stay-at-home orders impact a city's infection and fatality rates? How does the timing of the order influence these rates and how is hospital capacity effected?

#### For the project, I leverage an extension of the EpiModel package that was written by Tim Churches'. He has included many additional parameters, including hospitalization rates and hospital capacity which is one of the simulations I was interested in.   

load libraries :
```{r, message=FALSE, error=FALSE}
library(EpiModel)
library(extrafont)
library(animation)
library(tidyverse)
library(magrittr)
library(lubridate)
library(stringr)
library(tibble)
library(broom)
library(ggplot2)
#library(gt)
library(knitr)
library(devtools)
library(DiagrammeR)
library(parallel)
library(foreach)
library(tictoc)
suppressMessages(library(EpiModel))
library(incidence)
library(earlyR)
library(future)

#monkey patch the EpiModel extension

source_files <- c("_icm.mod.init.seiqhrf.R", "_icm.mod.status.seiqhrf.R", 
    "_icm.mod.vital.seiqhrf.R", "_icm.control.seiqhrf.R", "_icm.utils.seiqhrf.R", 
    "_icm.saveout.seiqhrf.R", "_icm.icm.seiqhrf.R")

src_path <- paste0("./_posts/2020-03-18-modelling-the-effects-of-public-health-", 
    "interventions-on-covid-19-transmission-part-2/")

gist_url <- "https://gist.github.com/timchurches/92073d0ea75cfbd387f91f7c6e624bd7"

local_source <- FALSE

for (source_file in source_files) {
    if (local_source) {
        source(paste(src_path, source_file, sep = ""))
    } else {
        source_gist(gist_url, filename = source_file)
    }
}
```

read in data :
*I chose a hospital capacity dataset from Data World. The numbers were provided by the Harvard Global Health Institute and are broken out by Hospital Referral Regions (HRR) for 305 local hospital markets in the country. At this granular level, it's possible to show how a region's resources may be stressed as COVID-19 infection rates rise.* 
```{r,warning=FALSE}
setwd("")
#read in and remove note on first line of csv
df <- read.csv("HRR_Population.csv",header = TRUE,colClasses = "character")[-1,]
attach(df)
#include population columns only, will do own projection calculations
df <- df[1:9]
head(df)
```

clean hospital data :
```{r}
#create city, state fields
#replace any space in City name with an underscore
hospital <- df%>%separate(HRR,c("City","State"),sep=",\\s*")

#change character values to numeric
hospital[,3:10] <- lapply(hospital[,3:10],function(hospital){as.numeric(gsub(",", "", hospital))})
head(hospital)
```

*To note, early in the project I looked to incorporate state obesity levels as it's been linked to patients with more severe cases. I wasn't clear on how to incorporate this into the model efficiently. But I left the data in the Appendix for future analysis.* 

```{r}
#total beds available
hospital <- hospital %>% mutate(Available_Beds_Total = (Available.Hospital.Beds + Available.ICU.Beds))
hospital <- hospital %>% mutate(Potential_Available_Beds_Total = Potentially.Available.Hospital.Beds. + Potentially.Available.ICU.Beds.)
```

# The EpiModel extension is named the Susceptible-Exposed-Infectious-Quarantined-Hospitalized-Recovered-Fatal model (`SEIQHRF`)

## Definitions :

* `Susceptible` = Adult population of the city
* `Exposed` = Exposed and infected, asymptomatic person who is potentially infectious to susceptibles
* `Infected` = Infected and symptomatic person who is infectious to susceptibles 
* `Quarantine` = An infected person self-isolates
* `Hospitalized` = Rate of progression from Infected to Hospitalized to Fatality
* `Recover` = Rate of progression from Hospitalized to discharge
* `Fatality` = Rate of progression from Hospitalized to death
* `Compartment` = Terminology used for the above terms in the Individual Compartmental Model (ICM)

Create a simulation function in order to change parameters efficiently :

This is the **baseline simulation** so the assumptions are essentially business as usual in the city. I start with 1 `infected` and a 2.5% probability of transmission to a `susceptible` (based on preliminary numbers from CDC). Of those that are `infected`, only a small portion of them `self-quarantine` and daily interactions remain high.

For my simulation cities, I chose **Boston, MA** and **Atlanta, GA**. I chose Boston because we live here, but also the timelines are very clear and detailed. I also wanted to simulate a city that wasn't following protocols and Columbus' recent protests prompted my choice. 

*Boston has an adult population >4M, so I scaled the numbers by 100 so I could run the simulations on my computer.* 

### City Simulation #1 : Boston, MA
```{r}
#test <- hospital_obs %>% filter(City =="Great Falls")#114,922
#test
boston <- hospital %>% filter(City=="Boston")
city1 <- boston
#scale
city1[3:12] <- city1[3:12]/100
city1
```


```{r}
# function to set-up and run the baseline simulations
# i kept many of the defaults here, but made some tweaks
simulate_bos <- function( #control
                     type = "SEIQHRF", 
                     nsteps = 366, #number of sim days, day 1 is initialization so 366 for full year
                     nsims = 5, #take average of n number of sims
                     ncores = 4, # cores to use
                     prog.rand = FALSE, #false for random draws from weibull distribution, random binomial draws were computationally difficult
                     rec.rand = FALSE, # random draws from weibull distribtuion at prog.rate for the progression from exposed + asymptomatic to infected + symptomatic
                     fat.rand = TRUE, #random draws from weibull distribution at fat.rate.base for case fatality progression from hospitalization to fatal case
                     quar.rand = FALSE, #random sample with sample faction at quar.rate for self-iso progression from infected + symptomatic to quarantine
                     hosp.rand = FALSE, #random sample with a sample fraction at hosp.rate for progression from infected + symptomatic or quarantine to hospitalization
                     disch.rand = TRUE, #random sample with sample fraction at disch.rate for progression from hospitalization to discharge
                     infection.FUN = infection.seiqhrf.icm, #default function for infection process
                     recovery.FUN = progress.seiqhrf.icm, #default function for recovery process
                     departures.FUN = departures.seiqhrf.icm, #default function for non-COVID deaths
                     arrivals.FUN = arrivals.icm, #default function for arrivals (births)
                     get_prev.FUN = get_prev.seiqhrf.icm, #default utility function to store values
                     
                     # initial 
                     s.num=city1$Adult.Population, #initial population of susceptible persons
                     e.num=0, #initial number of exposed + asymptomatic
                     i.num = 1, #initial number of infected + symptomatic
                     q.num=0, # initial number of quarantined population
                     h.num=0, #initial number of COVID needing hospitalization
                     r.num = 0, #intial number of COVID recovered
                     f.num = 0, #initial number of COVID deaths
                     
                     # params
                     inf.prob.e = 0.025, #prob. of an exposed + asymp person infecting a susceptible person 
                     act.rate.e = 10, # number of actions between exposed + asymp and a sus. person per day
                     inf.prob.i = 0.05, # prob. of infected person infecting a sus. person 
                     act.rate.i = 10, #number of actions between infected + symp person and a sus. person per day
                     inf.prob.q = 0.02, #prob. of infected quarantine person infecting a sus. person
                     act.rate.q = 2.5, # number of actions between quarantine person and a sus. person per day                    
                     quar.rate = 1/30, # rate per day an infected person enters quarantine, very low rate where most people are not self-quarantine
                     hosp.rate = 1/100, #rate per day an infected or quarantine person needs to be hospitalized 1% per day
                     disch.rate = 1/15, #rate per day of discharges from hospital
                     prog.rate = 1/10, #rate per day an exposed person becomes an infected person
                     prog.dist.scale = 5,
                     prog.dist.shape = 1.5,
                     rec.rate = 1/20,# rate per day an infected person recovers
                     rec.dist.scale = 35, #
                     rec.dist.shape = 1.5,
                     fat.rate.base = 1/50, #1/50 baseline COVID mortality rate per day for those needing hospitalization
                     hosp.cap = city1$Available_Beds_Total, #number of available hospital beds
                     fat.rate.overcap = 1/25, #mortality rate per day for people needing beds but who can't get one
                     fat.tcoeff = 0.5, #time coeff. for increasing mortality rate as a hospitalized person spends more time in the hospital
                     vital = TRUE, # enables arrivals and departures 
                     a.rate = 0, # arrival rate into the population
                     a.prop.e = 0.01,
                     a.prop.i = 0.001,
                     a.prop.q = 0.01,
                     ds.rate = (7/365)/1000, #non-COVID deaths
                     de.rate = (7/365)/1000, 
                     di.rate = (7/365)/1000, #non-COVID exit rate for infected
                     dq.rate = (7/365)/1000,
                     dh.rate = (20/365)/1000,
                     dr.rate = (7/365)/1000, #non-COVID exit rate for recovered 
                     out="mean" #summary function for simulations - take average
                    ) {
  control <- control.icm(type = type, 
                         nsteps = nsteps, 
                         nsims = nsims,
                         ncores = ncores,
                         prog.rand = prog.rand,
                         rec.rand = rec.rand,
                         infection.FUN = infection.FUN,
                         recovery.FUN = recovery.FUN,
                         arrivals.FUN = arrivals.FUN,
                         departures.FUN = departures.FUN,
                         get_prev.FUN = get_prev.FUN)

  init <- init.icm(s.num = s.num,
                   e.num = e.num,
                   i.num = i.num,
                   q.num = q.num,
                   h.num = h.num,
                   r.num = r.num,
                   f.num = f.num)

  param <-  param.icm(inf.prob.e = inf.prob.e, 
                      act.rate.e = act.rate.e,
                      inf.prob.i = inf.prob.i, 
                      act.rate.i = act.rate.i,
                      inf.prob.q = inf.prob.q, 
                      act.rate.q = act.rate.q,                    
                      quar.rate = quar.rate,
                      hosp.rate = hosp.rate,
                      disch.rate = disch.rate,
                      prog.rate = prog.rate,
                      prog.dist.scale = prog.dist.scale,
                      prog.dist.shape = prog.dist.shape,
                      rec.rate = rec.rate,
                      rec.dist.scale = rec.dist.scale,
                      rec.dist.shape = rec.dist.shape,
                      fat.rate.base = fat.rate.base,
                      hosp.cap = hosp.cap,
                      fat.rate.overcap = fat.rate.overcap,
                      fat.tcoeff = fat.tcoeff,
                      vital = vital,
                      a.rate = a.rate, 
                      a.prop.e = a.prop.e,
                      a.prop.i = a.prop.i,
                      a.prop.q = a.prop.q,
                      ds.rate = ds.rate, 
                      de.rate = de.rate, 
                      di.rate = di.rate,
                      dq.rate = dq.rate,
                      dh.rate = dh.rate,
                      dr.rate = dr.rate)

  sim <- icm.seiqhrf(param, init, control)
  sim_df <- as.data.frame(sim, out=out)

  return(list(sim=sim, df=sim_df))
}
set.seed(123)
baseline_sim <- simulate_bos(ncores=4)
```

Check the duration distributions from the baseline simulation:
```{r}
# function from developer that allows user to extract times and look at duration periods 
get_times <- function(simulate_results) {
    
    sim <- simulate_results$sim
    
    for (s in 1:sim$control$nsims) {
        if (s == 1) {
            times <- sim$times[[paste("sim", s, sep = "")]]
            times <- times %>% mutate(s = s)
        } else {
            times <- times %>% bind_rows(sim$times[[paste("sim", 
                s, sep = "")]] %>% mutate(s = s))
        }
    }
    
    times <- times %>% mutate(infTime = ifelse(infTime < 0, -5, 
        infTime), expTime = ifelse(expTime < 0, -5, expTime)) %>% 
        mutate(incubation_period = infTime - expTime, illness_duration = recovTime - 
            expTime, illness_duration_hosp = dischTime - expTime, 
            hosp_los = dischTime - hospTime, quarantine_delay = quarTime - 
                infTime, survival_time = fatTime - infTime) %>% 
        select(s, incubation_period, quarantine_delay, illness_duration, 
            illness_duration_hosp, hosp_los, survival_time) %>% 
        pivot_longer(-s, names_to = "period_type", values_to = "duration") %>% 
        mutate(period_type = factor(period_type, levels = c("incubation_period", 
            "quarantine_delay", "illness_duration", "illness_duration_hosp", 
            "hosp_los", "survival_time"), labels = c("Incubation period", 
            "Delay entering isolation", "Illness duration", "Illness (hosp)", 
            "Care required (hosp)", "Survival time of case fatalities"), 
            ordered = TRUE))
    return(times)
}
times <- get_times(baseline_sim)

times %>% filter(duration <= 30) %>% ggplot(aes(x = duration)) +
  geom_bar(fill="darkblue") + facet_grid(period_type ~ ., scales = "free_y") +theme(strip.text.y = element_text(size=16))+
 labs(title = "Boston Baseline Simulation Distributions") + theme(axis.text = element_text(size=16))
```

From the distribution plots, I may need to make some adjustments to the simulation. The reported incubation period for COVID is a little longer and the hospital care seems a little too skewed right. But the illness durations seem to make sense. 

Now that the **baseline simulation** for Boston has run successfully, we can plot the curves for the individual `Compartments` (definition above)
```{r}
baseline_plot_df <- baseline_sim$df %>% # use only the person columns
select(time, s.num, e.num, i.num, q.num, h.num, r.num, f.num) %>% 
#first 120 days
filter(time <= 120) %>% pivot_longer(-c(time), names_to = "Compartment", #individual compartmental model (ICM) 
    values_to = "count")

# colors to represent compartments
compcols <- c(s.num = "lightblue", e.num = "goldenrod1", i.num = "orangered1", 
    q.num = "cyan", h.num = "purple", r.num = "lightgreen", 
    f.num = "darkgray")
complabels <- c(s.num = "Susceptible", e.num = "Exposed", 
    i.num = "Infected", q.num = "Quarantine", h.num = "Hospitalized", 
    r.num = "Recovered", f.num = "Fatality")

baseline_plot_df %>% ggplot(aes(x = time, y = count, colour = `Compartment`)) + 
    geom_line(size = 1.2, alpha = 0.7) + scale_colour_manual(values = compcols, 
    labels = complabels) + theme(plot.background = element_rect(fill="gray")) + labs(title = "Boston Baseline Simulations (low rate of self-quarantine, social distance)", 
    x = "Days Since Start of COVID-19", y = "Persons (scaled 100)")


baseline_plot_df %>% filter(Compartment %in% c("e.num", "i.num", 
    "q.num", "h.num", "f.num")) %>% ggplot(aes(x = time, y = count, 
    colour = Compartment)) + geom_line(size = 1.2, alpha = 0.7) + 
    scale_colour_manual(values = compcols, labels = complabels) + 
    theme(plot.background = element_rect(fill="gray")) + labs(title = "BOS Baseline Simulations (low rate of self-quarantine, social distance)", x = "Days Since Start of COVID-19", 
    y = "Persons (scaled 100)")
--
```

Take a closer look at the distribution of the number of cases over time. 
```{r}
# incidence rate
case_counts <- baseline_sim$df %>% select(time, se.flow)
# uncount them
case_dates <- case_counts %>% uncount(se.flow) %>% 
    pull(time)
# convert to an incidence object
case_all <- case_dates %>% incidence(.)
# find the peak of the epidemic curve
peak_of_epidemic_curve <- find_peak(case_all)
peak_of_epidemic_curve
# plot the case curve

plot(case_all,xlab="Time",ylab = "Daily Case Count",color = "darkblue")
```

The **baseline simulation** indicates that the peak of the case curve is around day 40 and doesn't exactly "flatten", but start to fall off and COVID is over in 2 months.  

## From the plots, using the 'business as usual' response to COVID-19, simulates a pretty bleak outlook. There is a surge in `exposed` followed by `infected` around day 40 where about 1/3 of the population is infected (almost 1.5M). A small proporation `self-quarantine`. But, this is without any intervention and residents are not taking precautions. Now that we have the baseline simulation and some plots, we can start to see how government intervention (social distancing, stay-at-home orders) impact the baseline figures (for the better!). 

# [information as of 4/18] In Boston, the first confirmed case of COVID-19 was on February 1. Gov. Baker declared a state of emergency over a month later on March 10. By March 12, over 100 people had tested positive for the virus, in which most early cases were tracebale to the Biogen conference held February 26-27. On March 24, Baker announced a stay-at-home adivory effective until April 7 where non-essential businesses were ordered to close and restaurants and bars were restricted to takeout and delivery. Shortly after the closure was extended to May 4th. 

### Intervention #1 :
## Majority of `Infected` self-isolate. We can gradually increase the self-quarantine rate from 1/30 to 7/10 over two weeks. 
Starting March 2, the date of the second confirmed case, we can adjult the model to gradually increase the rate to 7/10 over the span of 2 weeks. I think it is a reasonable assumption that the majority of people did not begin self-isolation procedures until after the second confirmed case and that it would happen gradually. 

**(move Following Baker - Stay-at-home order starting 53 days after the first confirmed case. I'll increase both social distancing and increased self-isolation at that point, as there wasn't much in place and I was still commuting to work with most of Boston.)**

At <30 days, the self-isolation rate in the model is very low (1/30), but we can adjust it to gradually increase from day 30-45 up to 7/10.
```{r}
quar_rate_inc <- function(t) {
    ifelse(t < 15, 0.0333, ifelse(t <= 30, 0.0333 + (t - 15) * 
        (0.7 - 0.0333)/15, 0.70))
}
inc_quar_rate_sim <- simulate_bos(quar.rate = quar_rate_inc(1:366))
```

Now we can plot the new `Compartment` curves and see the improvement in outlook.
```{r}
inc_quar_rate_sim_plot_df <- inc_quar_rate_sim$df %>%
  # use only the persons columns
  select(time, s.num, e.num, i.num, q.num, 
         h.num, r.num, f.num) %>%
  # examine only the first 120 days
  filter(time <= 120) %>%
  pivot_longer(-c(time),
               names_to="Compartment",
               values_to="count")
inc_quar_rate_sim_plot_df %>%
  filter(Compartment %in% c("e.num","i.num",
                            "q.num","h.num",
                            "f.num")) #%>%
  #ggplot(aes(x=time, y=count/1000000, colour=Compartment)) +
    #geom_line(size=1.2, alpha=0.7) +
   # scale_colour_manual(values = compcols, labels=complabels) +
   # theme(plot.background = element_rect(fill="gray")) +
   # labs(title="Gradually increase self-isolation rate from day 30 to day 45",
    #     x="Days Since Start of COVID-19",
    #     y="Persons (millions)")


baseline_plot_df %>%
  mutate(experiment="Baseline") %>%
  bind_rows(inc_quar_rate_sim_plot_df %>%
              mutate(experiment="Gradual inc. self-isolation")) %>%
  filter(Compartment %in% c("e.num","i.num",
                            "q.num","h.num",
                            "f.num")) %>%
  ggplot(aes(x=time, y=count, colour=Compartment)) +
    geom_line(size=1.2, alpha=0.7) +
    facet_grid(experiment ~ .) +
    scale_colour_manual(values = compcols, labels=complabels) +
    theme(plot.background = element_rect(fill="gray")) +
    labs(title="BOS Simulation of baseline vs. gradual increase in self-isolation",
         x="Days Since Start of COVID-19",
         y="Persons (scaled 100)")

inc_quar_rate_sim_plot_df %>% filter(Compartment %in% c("e.num", "i.num"
    )) %>% ggplot(aes(x = time, y = count, 
    colour = Compartment)) + geom_line(size = 1.2, alpha = 0.7) + 
    scale_colour_manual(values = compcols, labels = complabels) + 
    theme(plot.background = element_rect(fill="gray")) + labs(title = "BOS Simulation of baseline vs. gradual increase in self-isolation)", x = "Days Since Start of COVID-19", 
    y = "Persons (scaled 100)")
inc_quar_rate_sim_plot_df
```

From the plot, it's clear that that increasing the self-isolation rate significantly impacts the infection curves (not too surprising) and the curve is flattening rather than a steep drop off.  

I can pull out the detailed difference between the hospitalization and fatality curves when comparing baseline with **intervention #1**. 

```{r}
baseline_plot_df %>%
  mutate(experiment="Baseline") %>%
  bind_rows(inc_quar_rate_sim_plot_df %>%
              mutate(experiment="Gradual inc. self-isolation")) %>%
  filter(Compartment %in% c("h.num",
                            "f.num")) %>%
  ggplot(aes(x=time, y=count, colour=Compartment)) +
    geom_line(size=1.2, alpha=0.7) +
  #change hospital count per use case
    geom_hline(yintercept = city1$Available_Beds_Total, colour="red", alpha=1) + 
    facet_grid(experiment ~ .) +
    scale_colour_manual(values = compcols, labels=complabels) +
    theme(plot.background = element_rect(fill="gray"))
    labs(title="Baseline vs gradual self-isolation simulations",
         x="Days Since Start of COVID-19",
         y="Persons")
```

The red line indicates hospital capacity (total available hospital and ICU beds).

From the plots, self-isolation alone has significantly reduced the strain on hospital capacity and reduced the fatality rate. The **baseline simulation** shows that without any self-isolation protocols, hospitals in Boston would have greatly exceeded capacity at the peak. But just gradually increasing self-isolation after becoming infected is not enough and hospital capacity is exceeded for an extended period of time. 


### Intervention #2 : 
## The *HRR dataset* includes potentially available bed capacity. The metric assumes that a potential bed becomes available if 50% of used beds could free up. Intervention #2 will take into account the gradual increase in self-isolation plus a gradual increase in bed capacity. 
```{r}
quar_rate_inc <- function(t) {
    ifelse(t < 15, 0.0333, ifelse(t <= 30, 0.0333 + (t - 15) * 
        (0.7 - 0.0333)/15, 0.7))
}
quar_rate_inc_sim <- simulate_bos(quar.rate = quar_rate_inc(1:366)) 

hosp_cap_inc <- function(t) {
    ifelse(t < 15, city1$Available_Beds_Total, ifelse(t <= 45, city1$Available_Beds_Total + (t - 15) * (city1$Potential_Available_Beds_Total - 
        city1$Available_Beds_Total)/30, city1$Potential_Available_Beds_Total))
}
#test same output as experiment #1 to make sure simulation runs properly
#inc_hosp_cap_sim <- simulate_bos(quar.rate = quar_rate_inc(1:366),hosp.cap = city1$Available_Beds_Total,s.num=city1$Adult.Population)
inc_hosp_cap_sim <- simulate_bos(quar.rate = quar_rate_inc(1:366),hosp.cap = hosp_cap_inc(1:366)) 

```

```{r}
inc_hosp_cap_sim_plot_df <- inc_hosp_cap_sim$df %>%
  # use only the persons columns
  select(time, s.num, e.num, i.num, q.num, 
         h.num, r.num, f.num) %>%
  # examine only the first 120 days
  filter(time <= 120) %>%
  pivot_longer(-c(time),
               names_to="Compartment",
               values_to="count")
inc_hosp_cap_sim_plot_df %>%
  filter(Compartment %in% c("e.num","i.num",
                            "q.num","h.num",
                            "f.num")) #%>%
  #ggplot(aes(x=time, y=count/1000000, colour=Compartment)) +
    #geom_line(size=1.2, alpha=0.7) +
   # scale_colour_manual(values = compcols, labels=complabels) +
   # theme(plot.background = element_rect(fill="gray")) +
   # labs(title="Gradually increase self-isolation rate from day 30 to day 45",
    #     x="Days Since Start of COVID-19",
    #     y="Persons (millions)")


baseline_plot_df %>%
  mutate(experiment="Baseline") %>%
  bind_rows(inc_hosp_cap_sim_plot_df %>%
              mutate(experiment="Gradual inc. self-quarantine, hospital capacity")) %>%
  filter(Compartment %in% c("e.num","i.num",
                            "q.num","h.num",
                            "f.num")) %>%
  ggplot(aes(x=time, y=count, colour=Compartment)) +
    geom_line(size=1.2, alpha=0.7) +
     facet_grid(experiment ~ .,labeller = label_wrap_gen(width = 35,multi_line = TRUE)) +
    scale_colour_manual(values = compcols, labels=complabels) +
    theme(plot.background = element_rect(fill="gray")) +
    labs(title="BOS Baseline vs. gradual inc. self-quarantine, hospital capacity",
         x="Days Since Start of COVID-19",
         y="Persons (scaled 100)")
```

```{r}
baseline_plot_df %>%
  mutate(experiment="Baseline") %>%
  bind_rows(inc_hosp_cap_sim_plot_df %>%
              mutate(experiment="Gradual inc. self-quarantine, hospital capacity")) %>%
  filter(Compartment %in% c("h.num",
                            "f.num")) %>%
  ggplot(aes(x=time, y=count, colour=Compartment)) +
    geom_line(size=1.2, alpha=0.7) +
  #change hospital count per use case
    geom_hline(yintercept = city1$Potential_Available_Beds_Total, colour="red", alpha=1) + 
    facet_grid(experiment ~ .,labeller = label_wrap_gen(width = 35,multi_line = TRUE)) +
    scale_colour_manual(values = compcols, labels=complabels) +
    theme(plot.background = element_rect(fill="gray"))
    labs(title="Baseline vs Gradual inc. self-quarantine, hospital capacity",
         x="Days Since Start of COVID-19",
         y="Persons")

```

From the plot, we can see that an increase in self-quarantine as well as an increase in available hospital beds, greatly reduces the strain put on hospitals during the peak of the outbreak. 

Now let's see how an increase in social distancing impacts these curve...
###Intervention #3 :
```{r}
quar_rate_inc <- function(t) {
    ifelse(t < 15, 0.0333, ifelse(t <= 30, 0.0333 + (t - 15) * 
        (0.7 - 0.0333)/15, 0.7))
}
quar_rate_inc_sim <- simulate_bos(quar.rate = quar_rate_inc(1:366)) 

hosp_cap_inc <- function(t) {
    ifelse(t < 15, city1$Available_Beds_Total, ifelse(t <= 45, city1$Available_Beds_Total + (t - 15) * (city1$Potential_Available_Beds_Total - 
        city1$Available_Beds_Total)/30, city1$Potential_Available_Beds_Total))
}
social_dist_inc <- function(t) {
    ifelse(t < 30, 10, ifelse(t <= 45, 10 - (t - 30) * (10 - 
        5)/15, 5))
}

social_dist_inc_sim <- simulate_bos(act.rate.i = social_dist_inc(1:366), 
    act.rate.e = social_dist_inc(1:366),quar.rate = quar_rate_inc(1:366),hosp.cap = hosp_cap_inc(1:366))
```

```{r}
social_dist_inc_sim_plot_df <- social_dist_inc_sim$df %>%
  # use only the persons columns
  select(time, s.num, e.num, i.num, q.num, 
         h.num, r.num, f.num) %>%
  # examine only the first 120 days
  filter(time <= 120) %>%
  pivot_longer(-c(time),
               names_to="Compartment",
               values_to="count")
social_dist_inc_sim_plot_df %>%
  filter(Compartment %in% c("e.num","i.num",
                            "q.num","h.num",
                            "f.num")) #%>%
  #ggplot(aes(x=time, y=count/1000000, colour=Compartment)) +
    #geom_line(size=1.2, alpha=0.7) +
   # scale_colour_manual(values = compcols, labels=complabels) +
   # theme(plot.background = element_rect(fill="gray")) +
   # labs(title="Gradually increase self-isolation rate from day 30 to day 45",
    #     x="Days Since Start of COVID-19",
    #     y="Persons (millions)")


baseline_plot_df %>%
  mutate(experiment="Baseline") %>%
  bind_rows(inc_hosp_cap_sim_plot_df %>%
              mutate(experiment="Gradual inc. self-quarantine, social dist, hospital capacity")) %>%
  filter(Compartment %in% c("e.num","i.num",
                            "q.num","h.num",
                            "f.num")) %>%
  ggplot(aes(x=time, y=count, colour=Compartment)) +
    geom_line(size=1.2, alpha=0.7) +
    facet_grid(experiment ~ .,labeller = label_wrap_gen(width = 35,multi_line = TRUE)) +
    scale_colour_manual(values = compcols, labels=complabels) +
    theme(plot.background = element_rect(fill="gray")) +
    labs(title="BOS Baseline vs. gradual inc. self-quarantine, social dist, hospital capacity",
         x="Days Since Start of COVID-19",
         y="Persons (scaled 100)")
```

```{r}
baseline_plot_df %>%
  mutate(experiment="Baseline") %>%
  bind_rows(social_dist_inc_sim_plot_df %>%
              mutate(experiment="Gradual inc. self-quarantine, social dist, hospital capacity")) %>%
  filter(Compartment %in% c("h.num",
                            "f.num")) %>%
  ggplot(aes(x=time, y=count, colour=Compartment)) +
    geom_line(size=1.2, alpha=0.7) +
  #change hospital count per use case
    geom_hline(yintercept = city1$Potential_Available_Beds_Total, colour="red", alpha=0.5) + 
    facet_grid(experiment ~ .,labeller = label_wrap_gen(width = 35,multi_line = TRUE)) +
    scale_colour_manual(values = compcols, labels=complabels) +
    theme(plot.background = element_rect(fill="gray"))
    labs(title="Baseline vs Gradual inc. hospital capacity",
         x="Days Since Start of COVID-19",
         y="Persons")

```


##Compare the Interventions
```{r}
plot_df <- baseline_sim$df %>% select(time, s.num, e.num, i.num, 
    q.num, h.num, r.num, f.num) %>% mutate(experiment = "#0 Baseline") %>% 
    bind_rows(inc_quar_rate_sim$df %>% select(time, s.num, e.num, 
        i.num, q.num, h.num, r.num, f.num) %>% mutate(experiment = "#1")) %>% 
    bind_rows(inc_hosp_cap_sim$df %>% select(time, s.num, e.num, 
        i.num, q.num, h.num, r.num, f.num) %>% mutate(experiment = "#2")) %>% 
    bind_rows(social_dist_inc_sim$df %>% select(time, s.num, 
        e.num, i.num, q.num, h.num, r.num, f.num) %>% mutate(experiment = "#3")) %>% 
    filter(time <= 150) %>% pivot_longer(-c(time, experiment), 
    names_to = "compartment", values_to = "count") %>% filter(compartment %in% 
    c("e.num", "i.num", "q.num", "h.num", "f.num"))

plot_df %>% ggplot(aes(x = time, y = count, colour = compartment)) + 
    geom_line(size = 1.2, alpha = 0.7) + facet_grid(experiment ~ .,labeller = label_wrap_gen(width = 35,multi_line = TRUE)) + scale_colour_manual(values = compcols, labels = complabels) + 
    theme(plot.background = element_rect(fill="gray")) + labs(title = "Baseline vs experiments", x = "Days since start of COVID-19", 
    y = "Persons (scaled 100)")
```

```{r}
plot_df %>% filter(compartment %in% c("h.num", "f.num")) %>% 
    ggplot(aes(x = time, y = count, colour = compartment)) + 
    geom_line(size = 1.2, alpha = 0.7) + facet_grid(experiment ~ 
    .) + scale_colour_manual(values = compcols, labels = complabels) + 
    theme(plot.background = element_rect(fill="gray")) + labs(title = "Baseline vs experiments", x = "Days since start of COVID-19", 
    y = "Persons")
```

City Simulation #2 : Atlanta, GA
```{r}
atlanta <- hospital %>% filter(City =="Atlanta")
city2 <- atlanta
#scale
city2[3:12] <- city2[3:12]/100
city2
```

Same setup for baseline simulation expect for adult population and hospital bed numbers
```{r}
# function to set-up and run the baseline simulations
# i kept many of the defaults here, but made some tweaks
simulate_atl <- function( #control
                     type = "SEIQHRF", 
                     nsteps = 366, #number of sim days, day 1 is initialization so 366 for full year
                     nsims = 5, #take average of n number of sims
                     ncores = 4, # cores to use
                     prog.rand = FALSE, #false for random draws from weibull distribution, random binomial draws were computationally difficult
                     rec.rand = FALSE, # random draws from weibull distribtuion at prog.rate for the progression from exposed + asymptomatic to infected + symptomatic
                     fat.rand = TRUE, #random draws from weibull distribution at fat.rate.base for case fatality progression from hospitalization to fatal case
                     quar.rand = FALSE, #random sample with sample faction at quar.rate for self-iso progression from infected + symptomatic to quarantine
                     hosp.rand = FALSE, #random sample with a sample fraction at hosp.rate for progression from infected + symptomatic or quarantine to hospitalization
                     disch.rand = TRUE, #random sample with sample fraction at disch.rate for progression from hospitalization to discharge
                     infection.FUN = infection.seiqhrf.icm, #default function for infection process
                     recovery.FUN = progress.seiqhrf.icm, #default function for recovery process
                     departures.FUN = departures.seiqhrf.icm, #default function for non-COVID deaths
                     arrivals.FUN = arrivals.icm, #default function for arrivals (births)
                     get_prev.FUN = get_prev.seiqhrf.icm, #default utility function to store values
                     
                     # initial 
                     s.num=city2$Adult.Population, #initial population of susceptible persons
                     e.num=0, #initial number of exposed + asymptomatic
                     i.num = 1, #initial number of infected + symptomatic
                     q.num=0, # initial number of quarantined population
                     h.num=0, #initial number of COVID needing hospitalization
                     r.num = 0, #intial number of COVID recovered
                     f.num = 0, #initial number of COVID deaths
                     
                     # params
                     inf.prob.e = 0.025, #prob. of an exposed + asymp person infecting a susceptible person 
                     act.rate.e = 10, # number of actions between exposed + asymp and a sus. person per day
                     inf.prob.i = 0.05, # prob. of infected person infecting a sus. person 
                     act.rate.i = 10, #number of actions between infected + symp person and a sus. person per day
                     inf.prob.q = 0.02, #prob. of infected quarantine person infecting a sus. person
                     act.rate.q = 2.5, # number of actions between quarantine person and a sus. person per day                    
                     quar.rate = 1/30, # rate per day an infected person enters quarantine, very low rate where most people are not self-quarantine
                     hosp.rate = 1/100, #rate per day an infected or quarantine person needs to be hospitalized 1% per day
                     disch.rate = 1/15, #rate per day of discharges from hospital
                     prog.rate = 1/10, #rate per day an exposed person becomes an infected person
                     prog.dist.scale = 5,
                     prog.dist.shape = 1.5,
                     rec.rate = 1/20,# rate per day an infected person recovers
                     rec.dist.scale = 35, #
                     rec.dist.shape = 1.5,
                     fat.rate.base = 1/50, #1/50 baseline COVID mortality rate 
                     hosp.cap = city2$Available_Beds_Total, #number of available hospital beds
                     fat.rate.overcap = 1/25, #mortality rate per day for people needing beds but who can't get one
                     fat.tcoeff = 0.5, #time coeff. for increasing mortality rate as a hospitalized person spends more time in the hospital
                     vital = TRUE, # enables arrivals and departures 
                     a.rate = 0, # arrival rate into the population
                     a.prop.e = 0.01,
                     a.prop.i = 0.001,
                     a.prop.q = 0.01,
                     ds.rate = (7/365)/1000, #non-COVID deaths
                     de.rate = (7/365)/1000, 
                     di.rate = (7/365)/1000, #non-COVID exit rate for infected
                     dq.rate = (7/365)/1000,
                     dh.rate = (20/365)/1000,
                     dr.rate = (7/365)/1000, #non-COVID exit rate for recovered 
                     out="mean" #summary function for simulations - take average
                    ) {
  control <- control.icm(type = type, 
                         nsteps = nsteps, 
                         nsims = nsims,
                         ncores = ncores,
                         prog.rand = prog.rand,
                         rec.rand = rec.rand,
                         infection.FUN = infection.FUN,
                         recovery.FUN = recovery.FUN,
                         arrivals.FUN = arrivals.FUN,
                         departures.FUN = departures.FUN,
                         get_prev.FUN = get_prev.FUN)

  init <- init.icm(s.num = s.num,
                   e.num = e.num,
                   i.num = i.num,
                   q.num = q.num,
                   h.num = h.num,
                   r.num = r.num,
                   f.num = f.num)

  param <-  param.icm(inf.prob.e = inf.prob.e, 
                      act.rate.e = act.rate.e,
                      inf.prob.i = inf.prob.i, 
                      act.rate.i = act.rate.i,
                      inf.prob.q = inf.prob.q, 
                      act.rate.q = act.rate.q,                    
                      quar.rate = quar.rate,
                      hosp.rate = hosp.rate,
                      disch.rate = disch.rate,
                      prog.rate = prog.rate,
                      prog.dist.scale = prog.dist.scale,
                      prog.dist.shape = prog.dist.shape,
                      rec.rate = rec.rate,
                      rec.dist.scale = rec.dist.scale,
                      rec.dist.shape = rec.dist.shape,
                      fat.rate.base = fat.rate.base,
                      hosp.cap = hosp.cap,
                      fat.rate.overcap = fat.rate.overcap,
                      fat.tcoeff = fat.tcoeff,
                      vital = vital,
                      a.rate = a.rate, 
                      a.prop.e = a.prop.e,
                      a.prop.i = a.prop.i,
                      a.prop.q = a.prop.q,
                      ds.rate = ds.rate, 
                      de.rate = de.rate, 
                      di.rate = di.rate,
                      dq.rate = dq.rate,
                      dh.rate = dh.rate,
                      dr.rate = dr.rate)

  sim <- icm.seiqhrf(param, init, control)
  sim_df <- as.data.frame(sim, out=out)

  return(list(sim=sim, df=sim_df))
}
set.seed(123)
baseline_sim <- simulate_atl(ncores=4)
```

Check the duration distributions from the baseline simulation:
```{r}
# function from developer that allows user to extract times and look at duration periods 
get_times <- function(simulate_results) {
    
    sim <- simulate_results$sim
    
    for (s in 1:sim$control$nsims) {
        if (s == 1) {
            times <- sim$times[[paste("sim", s, sep = "")]]
            times <- times %>% mutate(s = s)
        } else {
            times <- times %>% bind_rows(sim$times[[paste("sim", 
                s, sep = "")]] %>% mutate(s = s))
        }
    }
    
    times <- times %>% mutate(infTime = ifelse(infTime < 0, -5, 
        infTime), expTime = ifelse(expTime < 0, -5, expTime)) %>% 
        mutate(incubation_period = infTime - expTime, illness_duration = recovTime - 
            expTime, illness_duration_hosp = dischTime - expTime, 
            hosp_los = dischTime - hospTime, quarantine_delay = quarTime - 
                infTime, survival_time = fatTime - infTime) %>% 
        select(s, incubation_period, quarantine_delay, illness_duration, 
            illness_duration_hosp, hosp_los, survival_time) %>% 
        pivot_longer(-s, names_to = "period_type", values_to = "duration") %>% 
        mutate(period_type = factor(period_type, levels = c("incubation_period", 
            "quarantine_delay", "illness_duration", "illness_duration_hosp", 
            "hosp_los", "survival_time"), labels = c("Incubation period", 
            "Delay entering isolation", "Illness duration", "Illness (hosp)", 
            "Care required (hosp)", "Survival time of case fatalities"), 
            ordered = TRUE))
    return(times)
}
times <- get_times(baseline_sim)

times %>% filter(duration <= 30) %>% ggplot(aes(x = duration)) +
  geom_bar(fill="darkblue") + facet_grid(period_type ~ ., scales = "free_y") +theme(strip.text.y = element_text(size=16))+
 labs(title = "Atlanta Baseline Simulation Distributions") + theme(axis.text = element_text(size=16))
```

From the distribution plots, I may need to make some adjustments to the simulation. The reported incubation period for COVID is a little longer and the hospital care seems a little too skewed right. But the illness durations seem to make sense. 

Now that the **baseline simulation** for Atlanta has run successfully, we can plot the curves for the individual `Compartments` (definition above)
```{r}
baseline_plot_df <- baseline_sim$df %>% # use only the person columns
select(time, s.num, e.num, i.num, q.num, h.num, r.num, f.num) %>% 
#first 120 days
filter(time <= 120) %>% pivot_longer(-c(time), names_to = "Compartment", #individual compartmental model (ICM) 
    values_to = "count")

# colors to represent compartments
compcols <- c(s.num = "lightblue", e.num = "goldenrod1", i.num = "orangered1", 
    q.num = "cyan", h.num = "purple", r.num = "lightgreen", 
    f.num = "darkgray")
complabels <- c(s.num = "Susceptible", e.num = "Exposed", 
    i.num = "Infected", q.num = "Quarantine", h.num = "Hospitalized", 
    r.num = "Recovered", f.num = "Fatality")

baseline_plot_df %>% ggplot(aes(x = time, y = count, colour = `Compartment`)) + 
    geom_line(size = 1.2, alpha = 0.7) + scale_colour_manual(values = compcols, 
    labels = complabels) + theme(plot.background = element_rect(fill="gray")) + labs(title = "Atlanta Baseline Simulations (low rate of self-quarantine, social distance)", 
    x = "Days Since Start of COVID-19", y = "Persons (scaled 100)")


baseline_plot_df %>% filter(Compartment %in% c("e.num", "i.num", 
    "q.num", "h.num", "f.num")) %>% ggplot(aes(x = time, y = count, 
    colour = Compartment)) + geom_line(size = 1.2, alpha = 0.7) + 
    scale_colour_manual(values = compcols, labels = complabels) + 
    theme(plot.background = element_rect(fill="gray")) + labs(title = "Atlanta Baseline Simulations (low rate of self-quarantine, social distance)", x = "Days Since Start of COVID-19", 
    y = "Persons (scaled 100)")
```

Take a closer look at the distribution of the number of cases over time. 
```{r}
# incidence rate
case_counts <- baseline_sim$df %>% select(time, se.flow)
# uncount them
case_dates <- case_counts %>% uncount(se.flow) %>% 
    pull(time)
# convert to an incidence object
case_all <- case_dates %>% incidence(.)
# find the peak of the epidemic curve
peak_of_epidemic_curve <- find_peak(case_all)
peak_of_epidemic_curve
# plot the case curve

plot(case_all,xlab="Time",ylab = "Daily Case Count",color = "darkblue")
```

At <30 days, the self-isolation rate in the model is very low (1/30), but we can adjust it to gradually increase from day 15-30 up to 7/10. Will still assume that people quarantine when they become sick.
```{r}
quar_rate_inc <- function(t) {
    ifelse(t < 15, 0.0333, ifelse(t <= 30, 0.0333 + (t - 15) * 
        (0.7 - 0.0333)/15, 0.7))
}
inc_quar_rate_sim <- simulate_atl(quar.rate = quar_rate_inc(1:366))
```

Now we can plot the new `Compartment` curves and see the improvement in outlook.
```{r}
inc_quar_rate_sim_plot_df <- inc_quar_rate_sim$df %>%
  # use only the persons columns
  select(time, s.num, e.num, i.num, q.num, 
         h.num, r.num, f.num) %>%
  # examine only the first 120 days
  filter(time <= 120) %>%
  pivot_longer(-c(time),
               names_to="Compartment",
               values_to="count")
inc_quar_rate_sim_plot_df %>%
  filter(Compartment %in% c("e.num","i.num",
                            "q.num","h.num",
                            "f.num")) #%>%
  #ggplot(aes(x=time, y=count/1000000, colour=Compartment)) +
    #geom_line(size=1.2, alpha=0.7) +
   # scale_colour_manual(values = compcols, labels=complabels) +
   # theme(plot.background = element_rect(fill="gray")) +
   # labs(title="Gradually increase self-isolation rate from day 30 to day 45",
    #     x="Days Since Start of COVID-19",
    #     y="Persons (millions)")


baseline_plot_df %>%
  mutate(experiment="Baseline") %>%
  bind_rows(inc_quar_rate_sim_plot_df %>%
              mutate(experiment="Gradual inc. self-isolation")) %>%
  filter(Compartment %in% c("e.num","i.num",
                            "q.num","h.num",
                            "f.num")) %>%
  ggplot(aes(x=time, y=count, colour=Compartment)) +
    geom_line(size=1.2, alpha=0.7) +
    facet_grid(experiment ~ .) +
    scale_colour_manual(values = compcols, labels=complabels) +
    theme(plot.background = element_rect(fill="gray")) +
    labs(title="ATL Simulation of baseline vs. gradual increase in self-isolation",
         x="Days Since Start of COVID-19",
         y="Persons (scaled 100)")

inc_quar_rate_sim_plot_df %>% filter(Compartment %in% c("e.num", "i.num"
    )) %>% ggplot(aes(x = time, y = count, 
    colour = Compartment)) + geom_line(size = 1.2, alpha = 0.7) + 
    scale_colour_manual(values = compcols, labels = complabels) + 
    theme(plot.background = element_rect(fill="gray")) + labs(title = "ATL Simulation of baseline vs. gradual increase in self-isolation)", x = "Days Since Start of COVID-19", 
    y = "Persons (scaled 100)")
#inc_quar_rate_sim_plot_df
```

From the plot, it's clear that that increasing the self-isolation rate significantly impacts the infection curves (not too surprising) and the curve is flattening rather than a steep drop off.  

I can pull out the detailed difference between the hospitalization and fatality curves when comparing baseline with **intervention #1**. 

```{r}
baseline_plot_df %>%
  mutate(experiment="Baseline") %>%
  bind_rows(inc_quar_rate_sim_plot_df %>%
              mutate(experiment="Gradual inc. self-isolation")) %>%
  filter(Compartment %in% c("h.num",
                            "f.num")) %>%
  ggplot(aes(x=time, y=count, colour=Compartment)) +
    geom_line(size=1.2, alpha=0.7) +
  #change hospital count per use case
    geom_hline(yintercept = city2$Available_Beds_Total, colour="red", alpha=1) + 
    facet_grid(experiment ~ .) +
    scale_colour_manual(values = compcols, labels=complabels) +
    theme(plot.background = element_rect(fill="gray"))
    labs(title="Baseline vs gradual self-isolation simulations",
         x="Days Since Start of COVID-19",
         y="Persons")
```

### Intervention #2 : 
## The *HRR dataset* includes potentially available bed capacity. The metric assumes that a potential bed becomes available if 50% of used beds could free up. Intervention #2 will take into account the gradual increase in self-isolation plus a gradual increase in bed capacity. 
```{r}
quar_rate_inc <- function(t) {
    ifelse(t < 15, 0.0333, ifelse(t <= 30, 0.0333 + (t - 15) * 
        (0.7 - 0.0333)/15, 0.7))
}

hosp_cap_inc <- function(t) {
    ifelse(t < 15, city2$Available_Beds_Total, ifelse(t <= 45, city2$Available_Beds_Total + (t - 15) * (city2$Potential_Available_Beds_Total - 
        city2$Available_Beds_Total)/30, city2$Potential_Available_Beds_Total))
}
inc_hosp_cap_sim <- simulate_atl(quar.rate = quar_rate_inc(1:366),hosp.cap = hosp_cap_inc(1:366)) 

```

```{r}
inc_hosp_cap_sim_plot_df <- inc_hosp_cap_sim$df %>%
  # use only the persons columns
  select(time, s.num, e.num, i.num, q.num, 
         h.num, r.num, f.num) %>%
  # examine only the first 120 days
  filter(time <= 120) %>%
  pivot_longer(-c(time),
               names_to="Compartment",
               values_to="count")
inc_hosp_cap_sim_plot_df %>%
  filter(Compartment %in% c("e.num","i.num",
                            "q.num","h.num",
                            "f.num")) #%>%
  #ggplot(aes(x=time, y=count/1000000, colour=Compartment)) +
    #geom_line(size=1.2, alpha=0.7) +
   # scale_colour_manual(values = compcols, labels=complabels) +
   # theme(plot.background = element_rect(fill="gray")) +
   # labs(title="Gradually increase self-isolation rate from day 30 to day 45",
    #     x="Days Since Start of COVID-19",
    #     y="Persons (millions)")


baseline_plot_df %>%
  mutate(experiment="Baseline") %>%
  bind_rows(inc_hosp_cap_sim_plot_df %>%
              mutate(experiment="Gradual inc. self-quarantine, hospital capacity")) %>%
  filter(Compartment %in% c("e.num","i.num",
                            "q.num","h.num",
                            "f.num")) %>%
  ggplot(aes(x=time, y=count, colour=Compartment)) +
    geom_line(size=1.2, alpha=0.7) +
     facet_grid(experiment ~ .,labeller = label_wrap_gen(width = 35,multi_line = TRUE)) +
    scale_colour_manual(values = compcols, labels=complabels) +
    theme(plot.background = element_rect(fill="gray")) +
    labs(title="ATL Baseline vs. gradual inc. self-quarantine, hospital capacity",
         x="Days Since Start of COVID-19",
         y="Persons (scaled 100)")
```

```{r}
baseline_plot_df %>%
  mutate(experiment="Baseline") %>%
  bind_rows(inc_hosp_cap_sim_plot_df %>%
              mutate(experiment="Gradual inc. self-quarantine, hospital capacity")) %>%
  filter(Compartment %in% c("h.num",
                            "f.num")) %>%
  ggplot(aes(x=time, y=count, colour=Compartment)) +
    geom_line(size=1.2, alpha=0.7) +
  #change hospital count per use case
    geom_hline(yintercept = city2$Potential_Available_Beds_Total, colour="red", alpha=1) + 
    facet_grid(experiment ~ .,labeller = label_wrap_gen(width = 35,multi_line = TRUE)) +
    scale_colour_manual(values = compcols, labels=complabels) +
    theme(plot.background = element_rect(fill="gray"))
    labs(title="Baseline vs Gradual inc. self-quarantine, hospital capacity",
         x="Days Since Start of COVID-19",
         y="Persons")

```

###Intervention #3 :
```{r}
quar_rate_inc <- function(t) {
    ifelse(t < 15, 0.0333, ifelse(t <= 30, 0.0333 + (t - 15) * 
        (0.7 - 0.0333)/15, 0.7))
}

hosp_cap_inc <- function(t) {
    ifelse(t < 15, city2$Available_Beds_Total, ifelse(t <= 45, city2$Available_Beds_Total + (t - 15) * (city2$Potential_Available_Beds_Total - 
        city2$Available_Beds_Total)/30, city2$Potential_Available_Beds_Total))
}
social_dist_inc <- function(t) {
    ifelse(t < 15, 10, ifelse(t <=30, 10 - (t - 15) * (10 - 
        7.5)/15, 7.5))
}

social_dist_inc_sim <- simulate_atl(act.rate.i = social_dist_inc(1:366), 
    act.rate.e = social_dist_inc(1:366),quar.rate = quar_rate_inc(1:366),hosp.cap = hosp_cap_inc(1:366))
```

```{r}
social_dist_inc_sim_plot_df <- social_dist_inc_sim$df %>%
  # use only the persons columns
  select(time, s.num, e.num, i.num, q.num, 
         h.num, r.num, f.num) %>%
  # examine only the first 120 days
  filter(time <= 120) %>%
  pivot_longer(-c(time),
               names_to="Compartment",
               values_to="count")
social_dist_inc_sim_plot_df %>%
  filter(Compartment %in% c("e.num","i.num",
                            "q.num","h.num",
                            "f.num")) #%>%
  #ggplot(aes(x=time, y=count/1000000, colour=Compartment)) +
    #geom_line(size=1.2, alpha=0.7) +
   # scale_colour_manual(values = compcols, labels=complabels) +
   # theme(plot.background = element_rect(fill="gray")) +
   # labs(title="Gradually increase self-isolation rate from day 30 to day 45",
    #     x="Days Since Start of COVID-19",
    #     y="Persons (millions)")


baseline_plot_df %>%
  mutate(experiment="Baseline") %>%
  bind_rows(inc_hosp_cap_sim_plot_df %>%
              mutate(experiment="Gradual inc. self-quarantine, social dist, hospital capacity")) %>%
  filter(Compartment %in% c("e.num","i.num",
                            "q.num","h.num",
                            "f.num")) %>%
  ggplot(aes(x=time, y=count, colour=Compartment)) +
    geom_line(size=1.2, alpha=0.7) +
    facet_grid(experiment ~ .,labeller = label_wrap_gen(width = 35,multi_line = TRUE)) +
    scale_colour_manual(values = compcols, labels=complabels) +
    theme(plot.background = element_rect(fill="gray")) +
    labs(title="ATL Baseline vs. gradual inc. self-quarantine, social dist, hospital capacity",
         x="Days Since Start of COVID-19",
         y="Persons (scaled 100)")
```

```{r}
baseline_plot_df %>%
  mutate(experiment="Baseline") %>%
  bind_rows(social_dist_inc_sim_plot_df %>%
              mutate(experiment="Gradual inc. self-quarantine, social dist, hospital capacity")) %>%
  filter(Compartment %in% c("h.num",
                            "f.num")) %>%
  ggplot(aes(x=time, y=count, colour=Compartment)) +
    geom_line(size=1.2, alpha=0.7) +
  #change hospital count per use case
    geom_hline(yintercept = city2$Potential_Available_Beds_Total, colour="red", alpha=0.5) + 
    facet_grid(experiment ~ .,labeller = label_wrap_gen(width = 35,multi_line = TRUE)) +
    scale_colour_manual(values = compcols, labels=complabels) +
    theme(plot.background = element_rect(fill="gray"))
    labs(title="Baseline vs Gradual inc. hospital capacity",
         x="Days Since Start of COVID-19",
         y="Persons")

```

###Intervention #4 :
```{r}
quar_rate_inc <- function(t) {
    ifelse(t < 15, 0.0333, ifelse(t <= 30, 0.0333 + (t - 15) * 
        (0.7 - 0.0333)/15, 0.7))
}

#quar_rate_inc <- function(t) {
 #   ifelse(t < 15, 0.0333, ifelse(t <= 30, 0.0333 + (t - 15) * 
  #      (0.7 - 0.0333)/15, 0.7))
#}

hosp_cap_inc <- function(t) {
    ifelse(t < 15, city2$Available_Beds_Total, ifelse(t <= 45, city2$Available_Beds_Total + (t - 15) * (city2$Potential_Available_Beds_Total - 
        city2$Available_Beds_Total)/30, city2$Potential_Available_Beds_Total))
}

social_dist_inc2 <- function(t) {
    ifelse(t < 15, 10, ifelse(t <= 30, 10 - (t - 15) * (10 - 
        7.5)/15, ifelse(t<=60, 7.5- (t-30) * (7.5 - 10)/30,10)))
}


social_dist_inc_sim2 <- simulate_atl(act.rate.i = social_dist_inc2(1:366), 
    act.rate.e = social_dist_inc2(1:366),quar.rate = quar_rate_inc(1:366),hosp.cap = hosp_cap_inc(1:366))
```

```{r}
social_dist_inc_sim_plot_df <- social_dist_inc_sim2$df %>%
  # use only the persons columns
  select(time, s.num, e.num, i.num, q.num, 
         h.num, r.num, f.num) %>%
  # examine only the first 120 days
  filter(time <= 120) %>%
  pivot_longer(-c(time),
               names_to="Compartment",
               values_to="count")
social_dist_inc_sim_plot_df %>%
  filter(Compartment %in% c("e.num","i.num",
                            "q.num","h.num",
                            "f.num")) #%>%
  #ggplot(aes(x=time, y=count/1000000, colour=Compartment)) +
    #geom_line(size=1.2, alpha=0.7) +
   # scale_colour_manual(values = compcols, labels=complabels) +
   # theme(plot.background = element_rect(fill="gray")) +
   # labs(title="Gradually increase self-isolation rate from day 30 to day 45",
    #     x="Days Since Start of COVID-19",
    #     y="Persons (millions)")


baseline_plot_df %>%
  mutate(experiment="Baseline") %>%
  bind_rows(inc_hosp_cap_sim_plot_df %>%
              mutate(experiment="Gradual inc. self-quarantine, social dist, hospital capacity")) %>%
  filter(Compartment %in% c("e.num","i.num",
                            "q.num","h.num",
                            "f.num")) %>%
  ggplot(aes(x=time, y=count, colour=Compartment)) +
    geom_line(size=1.2, alpha=0.7) +
    facet_grid(experiment ~ .,labeller = label_wrap_gen(width = 35,multi_line = TRUE)) +
    scale_colour_manual(values = compcols, labels=complabels) +
    theme(plot.background = element_rect(fill="gray")) +
    labs(title="Baseline vs. gradual inc. self-quarantine, social dist, hospital capacity",
         x="Days Since Start of COVID-19",
         y="Persons (scaled 100)")
```

```{r}
baseline_plot_df %>%
  mutate(experiment="Baseline") %>%
  bind_rows(social_dist_inc_sim_plot_df %>%
              mutate(experiment="Gradual inc. self-quarantine, social dist, hospital capacity")) %>%
  filter(Compartment %in% c("h.num",
                            "f.num")) %>%
  ggplot(aes(x=time, y=count, colour=Compartment)) +
    geom_line(size=1.2, alpha=0.7) +
  #change hospital count per use case
    geom_hline(yintercept = city2$Potential_Available_Beds_Total, colour="red", alpha=0.5) + 
    facet_grid(experiment ~ .,labeller = label_wrap_gen(width = 35,multi_line = TRUE)) +
    scale_colour_manual(values = compcols, labels=complabels) +
    theme(plot.background = element_rect(fill="gray"))
    labs(title="Baseline vs Gradual inc. hospital capacity",
         x="Days Since Start of COVID-19",
         y="Persons")

```
##Compare the Interventions
```{r}
plot_df <- baseline_sim$df %>% select(time, s.num, e.num, i.num, 
    q.num, h.num, r.num, f.num) %>% mutate(experiment = "#0 Baseline") %>% 
    bind_rows(inc_quar_rate_sim$df %>% select(time, s.num, e.num, 
        i.num, q.num, h.num, r.num, f.num) %>% mutate(experiment = "#1")) %>% 
    bind_rows(inc_hosp_cap_sim$df %>% select(time, s.num, e.num, 
        i.num, q.num, h.num, r.num, f.num) %>% mutate(experiment = "#2")) %>% 
    bind_rows(social_dist_inc_sim$df %>% select(time, s.num, 
        e.num, i.num, q.num, h.num, r.num, f.num) %>% mutate(experiment = "#3")) %>% 
  bind_rows(social_dist_inc_sim2$df %>% select(time, s.num, 
        e.num, i.num, q.num, h.num, r.num, f.num) %>% mutate(experiment = "#4")) %>% 
    filter(time <= 150) %>% pivot_longer(-c(time, experiment), 
    names_to = "compartment", values_to = "count") %>% filter(compartment %in% 
    c("e.num", "i.num", "q.num", "h.num", "f.num"))

plot_df %>% ggplot(aes(x = time, y = count, colour = compartment)) + 
    geom_line(size = 1.2, alpha = 0.7) + facet_grid(experiment ~ .,labeller = label_wrap_gen(width = 35,multi_line = TRUE)) + scale_colour_manual(values = compcols, labels = complabels) + 
    theme(plot.background = element_rect(fill="gray")) + labs(title = "Baseline vs experiments", x = "Days since start of COVID-19", 
    y = "Persons (scaled 100)")
```

```{r}
plot_df %>% filter(compartment %in% c("h.num", "f.num")) %>% 
    ggplot(aes(x = time, y = count, colour = compartment)) + 
    geom_line(size = 1.2, alpha = 0.7) + facet_grid(experiment ~ 
    .) + scale_colour_manual(values = compcols, labels = complabels) + 
    theme(plot.background = element_rect(fill="gray")) + labs(title = "Baseline vs experiments", x = "Days since start of COVID-19", 
    y = "Persons")
```


# Appendix
read in data :
*adult obesity data by state from americashealthrankings.org 2019 report , there is a bit of lag in the data (2017)
https://www.americashealthrankings.org/explore/annual/measure/Obesity/state/ALL
Obesity has been linked to a very high percent of severe COVID patients* 
```{r}
#read in
df2<- read.csv("2019_HealthData.csv",header = TRUE,colClasses = "character")
attach(df2)
#include obesity measures only
obesity <- df2 %>% filter(Measure.Name=="Obesity")
obesity <- obesity %>% select(State.Name,Value)

#add state abbreviations for join
obesity$State <- state.abb[match(obesity$State.Name,state.name)]
obesity <- obesity %>% select(State, Value)
obesity$Value <- as.numeric(as.character(obesity$Value))
#conver to percentage
obesity$Value <- (obesity$Value)/100
#change column name
colnames(obesity) <- c("State","Obesity_Rate")
```

metric creation :
According to the CDC COVID-NET (as of 4/17/20), the cumulative COVID-19 hospitalization rates since March 1 are 20.0 per 100,000 with the highest rates in 65+ population at 63.8 per 100,000. The CDC also states that 12% of adult patients had underlying conditions, 48% of which were obsese. To note, COVID-NET is limited in its scope (as of April 18), it includes data from just 14 states (CA, CO, CT, GA, MD, MN, NM, NY, OR, TN, IA, MI, OH, UT). This represents approximately 10% of the US population (~32M).

* `COVID Adults` : Adult population infected based on % infected parameter
* `COVID Adults 65`: Percentage of 65+ adult population * 'COVID Adults'
* `Obese Adults` : Adult population * state obesity rate
* `COVID Adults Obese` : Percentage of obese adult population * 'COVID Adults'
* `Hospitalized 65` : 'COVID Adults 65' * 65+ hospitalizaiton rate
* `Hospitalized Obese` : 'COVID Adults Obese' * obese hospitalization rate
* `Hospitalized Other` : 'COVID Aults' * hospitalization rate - 'Hospitalized Obese'
* `Hospitalized Total` : Total projected to be hospitalized
* `Available Beds Total` : Combined available hospital and ICU beds for simplicity (potentially available if 50% of used beds could free up)

```{r}
#metric percent infected
infectioned_perc <- .20
hospital <- hospital %>% mutate(COVID_Adults = round(Adult.Population*(infectioned_perc)))
hospital <- hospital %>% mutate(COVID_Adults65 = round(COVID_Adults*(Population.65./Adult.Population)))

#add in obesity figures of the State, which has been indicated as a factor for higher hospitalization and mortality rate
hospital_obs <- merge(hospital, obesity, by="State")
hospital_obs <- hospital_obs %>%mutate(Obese_Adults = round(Adult.Population*(Obesity_Rate)))
hospital_obs <- hospital_obs %>% mutate(COVID_AdultsObese = round(COVID_Adults*(Obese_Adults/Adult.Population)))

#so, of those adults with COVID-19, I'll assume a number of the population that will be hospitalized based on CDC figures
#hospitalization_rate65 <- (63.8/100000)
hospital_obs <- hospital_obs %>% mutate(Hospitalized_65 = round(COVID_Adults65/100000*63.8))
#hospitalization_rate <- (20.0/100000)
hospital_obs <- hospital_obs %>% mutate(Hospitalized_Obese = round((Obese_Adults/100000*20.0*.12*.48)))
hospital_obs <- hospital_obs %>% mutate(Hospitalized_Other = round((COVID_Adults/100000*20.0)-(Hospitalized_Obese)-Hospitalized_65))
hospital_obs <- hospital_obs %>% mutate(Hospitalized_Total = Hospitalized_65+Hospitalized_Obese+Hospitalized_Other)
```
